Collaboration in Ad Hoc Teamwork: \\ Ambiguous Tasks, Roles, and Communication\vspace{1em}

### 2016-01-31 paper/adhoccom.tex
integrating Manuel comments

\begin{abstract}
Autonomous agents are becoming increasingly present in our environments, and will increasingly have to collaborate with previously unseen agents. Creating autonomous agents capable of cooperating with previously unfamiliar teammates, known as ``ad hoc teamwork'', has been identified as an important challenge for multiagent systems. Previous research has assumed that either the task, the role of each agent, or the communication protocol among agents is known before the interaction begins. In this paper, we consider these three variables simultaneously and show how an ad hoc agent can fit into a new team while handling ambiguous tasks, roles, and communication protocols. We assume a known distribution of possible tasks, roles, and communication protocols. We present experimental results in the pursuit domain showing that an ad hoc agent can integrate itself into a team by quickly identifying the task, selecting its role on the team, and estimating the communication protocol being used; thus barely impacting the overall team performance compared to a pre-coordinated agent.
\end{abstract}

>>>>

\begin{abstract}
% Autonomous agents are becoming increasingly present in our environments, and will increasingly have to collaborate with previously unseen agents.
Creating autonomous agents capable of cooperating with previously unfamiliar teammates, known as ``ad hoc teamwork'', has been identified as an important challenge for multiagent systems. Previous research has assumed that either the task, the role of each agent, or the communication protocol among agents is known before the interaction begins. We consider these three variables simultaneously and show how an ad hoc agent can fit into a new team while handling ambiguous tasks, roles, and communication protocols. We assume a known distribution of possible tasks, roles, and communication protocols. We present experimental results in the pursuit domain showing that our ad hoc agent can join such a team without barely impacting the overall performance compared to a pre-coordinated agent.
\end{abstract}

### 2016-01-31 paper/domain.tex
integrating Manuel comments

To select their actions, all our agents use a two step process. They first assign rewards to states they would like to reach. Then, knowing the full dynamics of the environment, they follow the optimal policy computed using reinforcement learning dynamic programming methods \cite{sutton1998reinforcement}, here value iteration using a discount factor of $0.95$. An agent considers all other agents as static obstacles. %keep this detail for the long version if accepted: If an other agent is in a state whose associated reward is non zero, it is not considered as an obstacle. Indeed, if there is only one state with non zero reward, and if this state is occupied, the agent could never reach such state and the optimal policy would be random -- therefore not optimally progressing towards that state.

@@

If the prey position is known exactly, each predator will aim at only one state, i.e. only one state will have non zero reward for the planning. This corresponds to the situation in Figure~\ref{fig:role11}. The same reasoning can be extended for a probabilistic knowledge of the prey state: for each prey state is associated a target state (as described above), to which we assign as reward the probability of the prey being in the state considered. This is of particular importance for the case of partial observability presented next.

@@

\subsubsection*{Message encoding} If one agent sees the prey, it can broadcast the position of the prey -- informing all other predators. Therefore, as soon as one agent is in close range with the prey, all other agents are informed, resuming in the full observability case described previously. %This explains the similar performance of the partial observability case with communication as show in Figure~\ref{fig:cmpteam}. Only the cases when no agent sees the prey decrease the performance of such teams.

@@

For a full team, the probability map of the prey state will never be zero, merging the information from messages of different agents will always be coherent. As we will see in the next section, this might not be the case when the ad hoc agent tries to understand what is going on, which will be valuable to inferring the team communication system.% As described in previous subsection, given a probability map of prey state, each predator computes a reward map given its specific role in the team (N/S/E/W), and selects the optimal action to maximize this reward.

>>>>

To select their actions, all our agents use a two step process. They first assign rewards to states they would like to reach. Then, knowing the full dynamics of the environment, they follow the optimal policy computed using dynamic programming methods \cite{sutton1998reinforcement}, here value iteration using a discount factor of $0.95$. An agent considers all other agents as static obstacles. %keep this detail for the long version if accepted: If an other agent is in a state whose associated reward is non zero, it is not considered as an obstacle. Indeed, if there is only one state with non zero reward, and if this state is occupied, the agent could never reach such state and the optimal policy would be random -- therefore not optimally progressing towards that state.

@@

If the prey position is known exactly, each predator will aim at only one state, i.e. only one state will have non zero reward value for the planning. This corresponds to the situation in Figure~\ref{fig:role11}. The same reasoning can be extended for a probabilistic knowledge of the prey state: for each prey state is associated a target state (as described above), to which we assign as reward the probability of the prey being in the state considered. This is of particular importance for the case of partial observability presented next.

@@

\subsubsection*{Message encoding} If one agent sees the prey, it can broadcast the position of the prey -- informing all other predators. Therefore, as soon as one agent is in close range with the prey, all other agents are informed, becoming the full observability case described previously. %This explains the similar performance of the partial observability case with communication as show in Figure~\ref{fig:cmpteam}. Only the cases when no agent sees the prey decrease the performance of such teams.

@@

For a full team, the probability map of the prey state will never be uniform, i.e.  merging the information from messages of different agents will always be coherent. As we will see in the next section, this might not be the case when the ad hoc agent tries to understand what is going on by interpreting messages according to different hypotheses on the communication protocol. Observing discrepency between messages will thus be valuable to inferring the team communication system.% As described in previous subsection, given a probability map of prey state, each predator computes a reward map given its specific role in the team (N/S/E/W), and selects the optimal action to maximize this reward.

### 2016-01-31 paper/introduction.tex
integrating Manuel comments

A similar problem, where both the task and the communication are unknown, has already been investigated in human-machine interaction \cite{grizou2014interactive}. They consider only two agents, a teacher and a learner. The learning agent has access to less information than the scenario presented here -- it could not observe the actions of the other agent and could only act on its own towards the success of the task. However, the learner could observe non-symbolic communication signals from the other agent, whose meaning, i.e. its language, was initially unknown. It is yet unclear if their approach can be adapted to multi-agent scenarios.

>>>>

A similar problem, where both the task and the communication are unknown, has been investigated in human-machine interaction \cite{grizou2014interactive}. They consider only two agents, a teacher and a learner. Their learning agent can only act on its own towards the success of the task and can observe non-symbolic communication signals from the teaching agent, whose meaning is part of finite set but initially unknown.

% A similar problem, where both the task and the communication are unknown, has been investigated in human-machine interaction \cite{grizou2014interactive}. They consider only two agents, a teacher and a learner. The learning agent has access to less information than the scenario presented here -- it could not observe the actions of the other agent and could only act on its own towards the success of the task. However, the learner could observe non-symbolic communication signals from the other agent, whose meaning, i.e. its language, was initially unknown. It is yet unclear if their approach can be adapted to multi-agent scenarios.

### 2016-01-31 paper/method.tex
integrating Manuel comments

First, the agent can use the observation of other agents' state as described in Equation~\ref{eq:obsupdate}. In our pursuit domain, the ad hoc agent knows the state of the all predators, but, in the partial observability case, it has uncertainty about the prey position.
%
\begin{eqnarray}
p(d_h) &=& \sum_{s'_{prey}} \sum_{s_{prey}} p(d_h|S',S) p(s'_{prey}) p(s_{prey})  \label{eq:adhocobsupdate}
\end{eqnarray}
%
with $p(d_h|S',S)$ as expanded in Equation~\ref{eq:obsupdateexpanded}. In the case of full observability, the sum over all possible prey states disappears. In the case of partial observability, messages allow to reduce the uncertainty about the prey state. It is very helpful to narrow the computation of Equation~\ref{eq:adhocobsupdate}. We explicitly write the state of the prey as $s_{prey}$ in the following equations. $s^{obs}_{prey}$ represents the information the ad hoc agent has about the prey before integrating information from the messages. Equation~\ref{eq:stateupdate} unfolds as:
%
\begin{eqnarray}
\lefteqn{p(s_{prey}|M, s^{obs}_{prey}, S, d_h)} \nonumber \\  &=& \prod_i p(s_{prey}|m_{b_i}, s_{b_i}, s^{obs}_{prey}, \com_h)
\label{eq:preyestimate}
\end{eqnarray}
%
with $\com_h$ from $d_h$ and because agents' messages are independent.

The estimation of the coherence of agents' messages $p(M|S,d_h)$ from Equation~\ref{eq:messageupdate} is computationally costly because the prey position is not fully determined to the ad hoc agent. It only has access to a probability map of the prey state.
% \todo{weird sentence::: It should therefore go over all hypothesis prey states while accounting for the potential noise in communication}.
To speed up the learning, it approximates Equation~\ref{eq:messageupdate} by summing the values of the prey state probability map inferred from the decoding of the messages in Equation~\ref{eq:preyestimate}.
%
\begin{eqnarray}
p(M|S,d_h) &\approx& \sum_{s} p(s_{prey} = s|M, s^{obs}_{prey}, S, d_h)  \label{eq:approxmessupdate}
\end{eqnarray}
%
For example, if the map is full of zeros, the information decoded from predators' messages is not coherent and therefore the hypothesis can be discarded, i.e. $p(M|S,d_h) = 0$. The more the maps decoded from each agent overlap, the higher the probability.

>>>>

First, the agent can use the observation of other agents' state as described in Equation~\ref{eq:obsupdate}. In our pursuit domain, the ad hoc agent knows the state of all the predators, but, in the partial observability case, it has uncertainty about the prey position.
%
\begin{eqnarray}
p(d_h) &=& \sum_{s'_{prey}} \sum_{s_{prey}} p(d_h|S',S) p(s'_{prey}) p(s_{prey})  \label{eq:adhocobsupdate}
\end{eqnarray}
%
with $p(d_h|S',S)$ as expanded in Equation~\ref{eq:obsupdateexpanded}. In the case of full observability, the sum over all possible prey states disappears. In the case of partial observability, messages allow to reduce the uncertainty about the prey state. It is very helpful to narrow the computation of Equation~\ref{eq:adhocobsupdate}. We explicitly write the state of the prey as $s_{prey}$ in the following equations. $s^{obs}_{prey}$ represents the information the ad hoc agent has about the prey before integrating information from the messages. Equation~\ref{eq:stateupdate} unfolds as:
%
\begin{eqnarray}
\lefteqn{p(s_{prey}|M, s^{obs}_{prey}, S, d_h)} \nonumber \\  &=& \prod_i p(s_{prey}|m_{b_i}, s_{b_i}, s^{obs}_{prey}, \com_h)
\label{eq:preyestimate}
\end{eqnarray}
%
with $\com_h$ from $d_h$ and because agents' messages are independent.

The estimation of the coherence of agents' messages $p(M|S,d_h)$ from Equation~\ref{eq:messageupdate} is computationally costly because the prey position is not fully observable to the ad hoc agent. It can only rely on a probability map of the prey state, therefore requiring to update on all states weigthed by their respective probability.
% \todo{weird sentence::: It should therefore go over all hypothesis prey states while accounting for the potential noise in communication}.
To speed up the process, we approximates Equation~\ref{eq:messageupdate} by summing the values of the prey state probability map inferred from the decoding of the messages in Equation~\ref{eq:preyestimate}.
%
\begin{eqnarray}
p(M|S,d_h) &\approx& \sum_{s} p(s_{prey} = s|M, s^{obs}_{prey}, S, d_h)  \label{eq:approxmessupdate}
\end{eqnarray}
%
For example, if the map is full of zeros, the information decoded from predators' messages is not coherent and therefore the hypothesis can be discarded, i.e. $p(M|S,d_h) = 0$. The more the maps decoded from each agent overlap, the higher the probability.

### 2016-01-31 paper/problem.tex
integrating Manuel comments

To tackle this problem, we assume the agent has access to a bigger set $D = \{d_1, \ldots, d_{n_H}\}$, containing $n_H$ possible domains, from which is pulled the particular domain $d$ considered. We further consider that, for any given $d_h$, the ad hoc agent can predict, in a probabilistic way, the expected behavior and communication of the agents. Hence, our approach relies on computing the posterior probability of each hypothetical domain given the information available to the ad hoc agent, here the observation from states and messages of the other agents. The correct hypothesis will be the one that maximizes this probability:
%
\begin{eqnarray}
\argmax_h~~p(d_h|S',S,M)  \label{eq:problem}
\end{eqnarray}
%
where $S$ and $S'$ are the observed states and next states, and $M$ is the messages sent by each agent. We can decompose the previous equation using Bayes' rule and see that we have to compute two different components: first the probability of the next states given the initial states, the messages, and a domain hypothesis $p(S'|S,M,d_h)$, and then the probability of the messages themselves given states and a domain hypothesis $p(M|S,d_h)$. In the following subsections, we detail these components as well as how the ad hoc agent can plan its actions.% first how to use observations of the state of other agents, then how to exploit their messages, and finally

@@

Observing the behavior of all other agents is a valuable source of information. Given a hypothesis domain $d_h$, we can compute the probability of the next agent state $S'$ given the current agent state $S$. For each hypothesis, we create a Bayes' filter that accumulates the probability of each domain conditionally on the observation of the agent movements.
To do so, we must estimate the probability that each agent selected each available action. We then estimate the probability of the observed state given all possible combinations of agents' actions and the environment dynamics:
% We denote that the computational cost of such update rule rises quickly with the number of agents.
%
\begin{eqnarray}
p(d_h|S',S) &\propto& p(S'|S,d_h) p(d_h)
\label{eq:obsupdate}
\end{eqnarray}
%
with
%
\begin{eqnarray}
p(S'|S,d_h) = \prod_i \sum_j p(s'_{b_i}|s_{b_i},u_j,\env_h) p(u_j|s_{b_i}, S, d_h)
\label{eq:obsupdateexpanded}
\end{eqnarray}
%
where $\env_h$ is the environment in $d_h$ which includes the state transition model. And $p(u_j|s_{b_i}, S, d_h)$ is the model of agent $b_i$ action selection, which is based on all the components of $d_h$ and the current state of the domain $S$. Given the agents' roles, they actions are independent. % But this might not always be the case, as exemplified in the pursuit domain example of section~\ref{sec:domain}. To accommodate with thisThe equation above considers the case of full knowledge of the states. As this might not always be true, the update rule should also account for partial knowledge of states.
The equation becomes:
%
\begin{eqnarray}
p(d_h) &=& \sum_{S'} \sum_S p(d_h|S',S) p(S') p(S)
\label{eq:obsupdatestateprob}
\end{eqnarray}
%
which can be expanded as for Equation~\ref{eq:obsupdateexpanded}. For example, this uncertainty about the state of the world can be explained by partial observability from the ad hoc agent. Communication between agents can help refine its estimate of the current state.

@@

We now consider the action selection method for the ad hoc agent. Previous work considered ad hoc agents that solve the optimal teammate problem. Such agents know the model of their teammates and select their next action in order to improve maximally the team performance -- often resulting in better performance that the initial team \cite{barrett2011empirical}. The aim of this study is different; we want to advocate the fact that the ad hoc agent can work under less assumptions than before and be able to estimate more information about the new team. Hence, we isolate our algorithm from the planning aspects and test whether it can select between candidate domain. Therefore, in this work, the ad hoc agent simply tries to replace a missing agent in the team. To this end, the ad hoc agent will weight the policies for each domain hypotheses $d_h$ by the probability currently assigned to this configuration $p(d_h)$.
%
\begin{eqnarray}
p(u_a|M, S,d_h) = \sum_h p(u_a|M, S, d_h) p(d_h)
\label{eq:adhocpolicy}
\end{eqnarray}
%
With this planning strategy, once the correct hypothesis is identified, the ad hoc agent will mimic the default behavior of the agent it replaces. But the agent is likely to make sensible decision earlier as unrelevant hypotheses are discarded.

>>>>

To tackle this problem, we assume the agent has access to a bigger set $D = \{d_1, \ldots, d_{n_H}\}$, containing $n_H$ possible domains, from which is pulled the particular domain $d$ considered. We further consider that, for any given $d_h$, the ad hoc agent can predict, in a probabilistic way, the expected behavior and communication of the agents. Hence, our approach relies on computing the posterior probability of each hypothetical domain given the information available to the ad hoc agent, here the observation from states and messages of the other agents. The correct hypothesis will be the one that maximizes this probability:
%
\begin{eqnarray}
\argmax_h~~p(d_h|S',S,M)  \label{eq:problem}
\end{eqnarray}
%
where $S$ and $S'$ are the observed states and next states, and $M$ is the messages sent by each agent. At each step a new tuple $\{S', S, M\}$ is observed and the probabilities are updated. Following Bayes' rule, we have to compute two different components: first the probability of the observed next states given the initial states, the messages, and a domain hypothesis $p(S'|S,M,d_h)$, and then the probability of the messages themselves given the oberved states and a domain hypothesis $p(M|S,d_h)$. In the following subsections, we detail these components as well as how the ad hoc agent plans its actions.% first how to use observations of the state of other agents, then how to exploit their messages, and finally

@@

Observing the behavior of all other agents is a valuable source of information. Given a hypothesis domain $d_h$, we can compute the probability of the next agent state $S'$ given the current agent state $S$. For each hypothesis, we create a Bayes' filter that accumulates the probability of each domain conditionally on the observation of the agent movements.
To do so, we must estimate the probability that each agent selected each available action. We then estimate the probability of the observed state given all possible combinations of agents' actions and the environment dynamics:
% We denote that the computational cost of such update rule rises quickly with the number of agents.
%
\begin{eqnarray}
p(d_h|S',S) &\propto& p(S'|S,d_h) p(d_h)
\label{eq:obsupdate}
\end{eqnarray}
%
with
%
\begin{eqnarray}
p(S'|S,d_h) = \prod_i \sum_j p(s'_{b_i}|s_{b_i},u_j,\env_h) p(u_j|s_{b_i}, S, d_h)
\label{eq:obsupdateexpanded}
\end{eqnarray}
%
where $\env_h$ is the environment in $d_h$ which includes the state transition model. And $p(u_j|s_{b_i}, S, d_h)$ is the model of agent $b_i$ action selection, which is based on all the components of $d_h$ and the current state of the domain $S$. Given the agents' roles, their actions are independent. % But this might not always be the case, as exemplified in the pursuit domain example of section~\ref{sec:domain}. To accommodate with this

The equation above considers the case of full observability of the states. As this might not always be true (e.g. partial observability from the ad hoc agent in section~\ref{sec:com}), the update rule should also account for partial observability, represented by a discrete probabily distribution on $S$ and $S'$. The update becomes:
%
\begin{eqnarray}
p(d_h) &=& \sum_{S'} \sum_S p(d_h|S',S) p(S') p(S)
\label{eq:obsupdatestateprob}
\end{eqnarray}
%
which can be expanded as Equation~\ref{eq:obsupdate} is in Equation~\ref{eq:obsupdateexpanded}.

@@

We now consider the action selection method for the ad hoc agent. Previous work considered ad hoc agents that solve the optimal teammate problem. Such agents know the model of their teammates and select their next action in order to improve maximally the team performance -- often resulting in better performance than the initial team \cite{barrett2011empirical}. The aim of this study is different; we want to advocate the fact that the ad hoc agent can work under less assumptions than before and be able to estimate more information about the new team. Hence, we isolate our algorithm from the planning aspects and test whether it can select between candidate domains. Therefore, in this work, the ad hoc agent simply tries to replace a missing agent in the team. To this end, the ad hoc agent will weight the policies for each domain hypotheses $d_h$ by the probability currently assigned to this configuration $p(d_h)$.
%
\begin{eqnarray}
p(u_a|M, S) = \sum_h p(u_a|M, S, d_h) p(d_h)
\label{eq:adhocpolicy}
\end{eqnarray}
%
With this planning strategy, once the correct hypothesis is identified, the ad hoc agent will mimic the default behavior of the agent it replaces. But the agent is likely to make sensible decision earlier as irrelevant hypotheses are discarded.

### 2016-01-31 paper/results.tex
fixed anonymous thingy
integrating Manuel comments

For each experiment run we randomly create a domain set $D$, comprised of a set of 10 task hypotheses, 10 team configurations, and 10 communication protocols; resulting in 1000 domains. Among this set, one configuration was selected for the team but was unknown to the ad hoc agent. All the figures presented next display the mean and standard error of the variable considered. Standard errors are shown has a shaded area, but, given the high number of samples (1000 runs using the same random seed for all conditions), it is barely visible. Statistical results presented are two-sample t-test to determine if the average final scores of teams are equal.

The code to reproduce these results is available online at \textbf{[anonymous for blind review]}.

A team of agents is evaluated by its total reward accumulated in 200 steps, i.e. the number of time the prey was captured. After each capture of the prey, the predators and the prey position are randomly reassigned. The capture state, the team configuration, and the communication protocol do not change during the 200 steps.

>>>>

For each experiment run we randomly create a domain set $D$, comprised of a set of 10 task hypotheses, 10 team configurations, and 10 communication protocols; resulting in 1000 domains. Among this set, one configuration was selected for the team but was unknown to the ad hoc agent. All the figures presented next display the mean and standard error of the variable considered. Standard errors are shown as a shaded area, but, given the high number of samples (1000 runs using the same random seed for all conditions), it is barely visible. Statistical results presented are two-sample t-test to determine if the average final scores of teams are equal.

The code to reproduce these results is available online at \url{https://github.com/jgrizou/adhoc_com}.

A team of agents is evaluated by its total reward accumulated in 200 steps, i.e. the number of times the prey was captured. After each capture of the prey, the predators and the prey position are randomly reassigned. The capture state, the team configuration, and the communication protocol do not change during the 200 steps.
